---
format:
  revealjs:
    theme: night
editor: visual
---

## The Changing Risks of Exposure to Gun Violence in Chicago {background-image="img/stephan-cassara-KnAIsBuitGg-unsplash.jpg" background-opacity=0.3}

Charles C. Lanfear

Robert J. Sampson

David S. Kirk

Rebecca Bucci




# The {background-image="img/max-bender-yb7Yg3Rv7WA-unsplash.jpg" background-opacity=0.3}

![](img/phdcn_logo_white.svg)

::: footer
Sampson, Robert J., David S. Kirk, and Rebecca Bucci. 2022. "Cohort Profile: Project on Human Development in Chicago Neighborhoods and Its Additions (PHDCN+)." *Journal of Developmental and Life-Course Criminology* 8.  https://doi.org/10.1007/s40865-022-00203-0.
:::

## Goal

### Examine life-course change and cohort differentiation

* Describe exposure to violence (ETV), continuity and change

* Correlates of ETV

* Long-term consequences of ETV

## Timeline

```{r, dev = "ragg_png",  dev.args = list(bg = 'transparent')}
library(tidyverse)
library(ggforce)
library(ggtext)

df <- tribble(
  ~cohort, ~time, ~year, ~age, 
  0,  1, 1995,  0,
  0,  2, 2021, 25,
  3,  1, 1995,  3,
  3,  2, 2002, 11,
  6,  1, 1995,  6,
  6,  2, 2002, 14,
  9,  1, 1995, 9, 
  9,  2, 2021, 34,
  12, 1, 1995, 12,
  12, 2, 2021, 37,
  15, 1, 1995, 15,
  15, 2, 2021, 40,
  18, 1, 1995, 18,
  18, 2, 2002, 26,
) 

cs_df <- tibble(
  survey = rep(c("PHDCN-CS", "CCAHS"), each = 7),
  year   = rep(c(1995, 2002), each = 7),
  cohort = rep(seq(0,18, by = 3), length.out = 14)
)


ggplot(df, aes(x = year, y = cohort, group = cohort)) + 
  geom_line(size = 3, color = "white") +
  geom_richtext(aes(label = age), size = 5) +
  theme_minimal(base_size = 24) +
  labs(y = "Cohort Ages", x= NULL) +
  scale_x_continuous(breaks = seq(1995, 2021, by = 5), limits = c(1994,2022)) +
  scale_y_continuous(limits = c(-1,22)) +
  geom_mark_rect(data = cs_df, aes(group = year), color = "red", expand = unit(6, "mm")) +
  annotate("text", x = 1998.5, y = 21, label = "Community\nSurveys", color = "red", size = 5) +
  theme(panel.grid = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(color = "grey90"),
        text = element_text(family = "sans",  color = "white"),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA))
```

## Contexts of Violence

```{r, dev = "ragg_png",  dev.args = list(bg = 'transparent')}
load("./data/cohort_age_hom.RData")

cohort_age_hom %>%
  mutate(cohort = str_remove(cohort, "Cohort "),
         cohort = fct_reorder(cohort, as.numeric(cohort)),
         cohort = fct_rev(cohort)) %>%
  ggplot(aes(x = age, y = hom_rate, color = cohort)) + 
  geom_line() + 
  facet_grid(cohort~ ., switch = "y") +  
  xlim(0, 40) +
  scale_y_continuous(breaks = c(20, 30), 
                     position = "right", 
                     sec.axis = sec_axis(trans = ~.*0, name = "Cohort")) +
  annotate("rect", xmin = 13, xmax = 25, ymin = -Inf, ymax = Inf, fill = "white", alpha = 0.3) +
  theme_minimal(base_size = 24) +
  labs(x = "Cohort member age", y = "Homicide Rate") +
  theme(panel.grid = element_blank(),
        axis.text = element_text(color = "grey90"),
        axis.text.y.right =  element_blank(),
        axis.text.y.left  = element_blank(),
        strip.text = element_text(color = "grey90", angle = 90),
        legend.position = "none",
        text = element_text(family = "sans",  color = "white"),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA))
```

::: footer
Homicide rate ranges between 16 and 37 per 100,000 population
:::


## Approach

> Brief mention of interval censoring, use of Turnbull NPMLE, SPT Prop Hazards

## Covariates

* Race
* Sex
* Immigrant Generation
* Neighborhood Context in Childhood
   * Tract average Homicide rate
   * Tract concentrated disadvantage
   * NC collective efficacy