---
format:
  revealjs:
    theme: [night, ref/phdcn_style.scss]
    logo: img/phdcn_logo_white.svg
    incremental: true
    self-contained: true
editor: visual
mouse-wheel: true
history: false
---

```{r setup}
#| include: false
library(tidyverse)
library(ggforce)
library(ggtext)
library(showtext)
library(icenReg)
library(interval)
library(marginaleffects)
library(broom)
load("./data/survival_data.RData")
load("../../../Kirk Projects/phdcn_gun_violence_exposures/data/output/npt_cohort_models.RData")
load("../../../Kirk Projects/phdcn_gun_violence_exposures/data/output/npt_race_models.RData")
load("../../../Kirk Projects/phdcn_gun_violence_exposures/data/output/npt_sex_models.RData")
load("../../../Kirk Projects/phdcn_gun_violence_exposures/data/output/spt_base_models.RData")
source("../../../Kirk Projects/phdcn_gun_violence_exposures/syntax/project_functions.R")
knitr::opts_chunk$set(dev = "svg",  
                      dev.args = list(bg = 'transparent'),
                      message = FALSE,
                      echo = FALSE,
                      warning = FALSE,
                      fig.showtext = TRUE,
                      fig.retina = 1,
                      fig.align = "center")
plot_font <- "Open Sans"
get_age_percents <- function(x, .race = NULL, .cohort = NULL){
  if(is.null(.race) & is.null(.cohort)){
    exposure_ecdf <- ecdf(x$age_right)
    return(exposure_ecdf(1:40))
  } else if (is.character(.race)){
    exposure_ecdf <- ecdf(x$age_right[x$race == .race])
    return(exposure_ecdf(1:40))
  } else if (is.character(.cohort)){
    exposure_ecdf <- ecdf(x$age_right[x$cohort == .cohort])
    return(exposure_ecdf(1:40))
  } else {
    stop("If provided, race or cohort must be a character value")
  }
}
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
```

```{r}
#| include: false
#| cache: false

font_add_google(name = plot_font)

```

## Inequalities in Life-Course Exposure to Gun Violence {background-image="img/stephan-cassara-KnAIsBuitGg-unsplash.jpg" background-opacity="0.3"}

Charles C. Lanfear

Robert J. Sampson

David S. Kirk

Rebecca Bucci

# Homicide in Chicago

```{r}
load("../../../Kirk Projects/phdcn_neighborhoods/crime/data/derived/hom_rates_race_df.RData")
# cause_of_death_gun <- read_tsv("./data/cause_of_death_gun.tsv")
# cause_of_death_all <- read_tsv("./data/cause_of_death_all.tsv")
# cause_of_death_gun %>%
#   filter( (Notes == "Total" & `Hispanic Origin` == "Hispanic or Latino" ) | (`Hispanic Origin` == "Not Hispanic or Latino" & Race %in% c("White", "Black or African American"))) %>%
#   mutate(Race = ifelse(`Hispanic Origin` == "Hispanic or Latino", "Hispanic", str_c("NH ", str_remove(Race, " .*$"))),
#          `Rate per 1,000` = as.numeric(`Crude Rate`),
#          Type = "Gun") %>%
#   select(Year, Race, Type, `Rate per 1,000`) %>%
#   bind_rows(
#     cause_of_death_all %>%
#       filter( (Notes == "Total" & `Hispanic Origin` == "Hispanic or Latino" ) | 
#               (`Hispanic Origin` == "Not Hispanic or Latino" & Race %in% c("White", "Black or African American"))) %>%
#        mutate(Race = ifelse(`Hispanic Origin` == "Hispanic or Latino", "Hispanic", str_c("NH ", str_remove(Race, " .*$"))),
#               `Rate per 1,000` = as.numeric(`Crude Rate`),
#               Type = "All") %>%
#   select(Year, Race, Type, `Rate per 1,000`)) %>%
hom_rates_race_df %>%
  filter(race != "Other") %>%
  ggplot(aes(x = year, y = value, group = race, color = race)) + 
  geom_line(data = ~ . |> filter(type == "Gun Homicide"), linetype = "solid")  +
  geom_line(data = ~ . |> filter(type == "All Homicide"), linetype = "dashed")  +
  geom_richtext(data =  ~ . |> 
                  group_by(race) %>% 
                  filter(year == 1992) |>
                  summarize(value = mean(value), year = 1990.5), 
                aes(label = race), fill = "black") +
  geom_hline(yintercept = 0, color = "white", size = 0.25) +
  annotate("text", x = c(1990.5, 1990.5), y = c(10, 12), label = c("Gun Homicide", "All Homicide"), color = "white", family = plot_font) +
  geom_line(data = tibble(x = c(1993, 1994, 1993, 1994), 
                          y = c(10, 10, 12, 12), 
                          type = factor(c("Gun Homicide", "Gun Homicide", "All Homicide", "All Homicide"), levels = c("Gun Homicide", "All Homicide"))),
            aes(x = x, y = y, linetype = type), color = "white", family = plot_font, inherit.aes=FALSE) +
  scale_y_continuous(breaks = seq(0, 30, by = 10)) +
  coord_cartesian(clip = "off", ylim = c(0, 30), xlim = c(1990, 2021)) +
  xlab(NULL) + ylab("Rates per 100,000") +
  theme_minimal(base_size = 24) +
  theme(panel.grid = element_blank(),
        axis.text = element_text(color = "white"),
        text = element_text(family = plot_font,  color = "white"),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        legend.position = "none")
```

::: notes
Persistent large racial inequality in violent victimization

Guns used in an increasingly large proportion

Overall homicide rate in Chicago is 20 times any major European City
:::

## Questions {background-image="img/stephan-cassara-KnAIsBuitGg-unsplash.jpg" background-opacity="0.1"}

 

::: fragment
To what extent were individuals growing up in Chicago in this period exposed to **gun violence**?
:::

 

::: fragment
And how does this exposure differ...
:::

::: fragment
| ... by **race** and **sex**?
:::

::: fragment
| ... by **neighbourhood** context?
:::


::: notes
We'll be focusing mostly on race and sex here today. Next session will discuss cohorts.
:::

# The {background-image="img/max-bender-yb7Yg3Rv7WA-unsplash.jpg" background-opacity="0.3"}

![](img/phdcn_logo_white.svg)

## {background-image="img/max-bender-yb7Yg3Rv7WA-unsplash.jpg" background-opacity="0.1"}


### Project on Human Development in Chicago Neighborhoods

::: {.nonincremental}
-   6200 children in 7 birth cohorts
-   3 interviews from 1995--2002
-   Community surveys in 1995 and 2002
:::


## {background-image="img/max-bender-yb7Yg3Rv7WA-unsplash.jpg" background-opacity="0.1"}

### Project on Human Development in Chicago Neighborhoods

::: {.nonincremental}
-   6200 children in 7 birth cohorts
-   3 interviews from 1995--2002
-   Community surveys in 1995 and 2002
:::


### PHDCN+

::::: {.columns}

:::: {.column width="80%"}
::: {.nonincremental}
-   Representative sample of 4 cohorts
-   1057 interviewed in 2012
-   682 followed-up in 2021
:::
::::

:::: {.column width="20%"}

![](img/ncgvr.svg)

::::

:::::

::: footer
See Sampson, Kirk, & Bucci (2022)
:::



## Timeline

```{r}
#| fig.width: 9
#| fig.height: 6
df <- tribble(
  ~cohort, ~time, ~year, ~age, 
  0,  1, 1995,  0,
  0,  2, 2021, 25,
  3,  1, 1995,  3,
  3,  2, 2002, 11,
  6,  1, 1995,  6,
  6,  2, 2002, 14,
  9,  1, 1995, 9, 
  9,  2, 2021, 34,
  12, 1, 1995, 12,
  12, 2, 2021, 37,
  15, 1, 1995, 15,
  15, 2, 2021, 40,
  18, 1, 1995, 18,
  18, 2, 2002, 26,
) %>%
  mutate(focus = ifelse(cohort %in% c(0,9,12,15), "yes", "no"))

cs_df <- tibble(
  survey = rep(c("PHDCN-CS", "CCAHS"), each = 7),
  year   = rep(c(1995, 2002), each = 7),
  cohort = rep(seq(0,18, by = 3), length.out = 14)
)
wave_df <- tibble(
  survey = rep(1:5, length.out = 35),
  year   = rep(c(1995, 1998.5, 2002, 2012, 2021), length.out = 35),
  cohort = rep(seq(0,18, by = 3), each = 5)
) %>%
  filter(year <= 2002 | cohort %in% c(0, 9, 15)) %>%
  mutate(group = 
           case_when(
            year <= 2002 ~ year, 
            year > 2002 & cohort %in% c(9,15,18) ~ year,
            year > 2002 & cohort == 0 ~ year -1),
         phdcn = ifelse(year <= 2002, "PHDCN", "PHDCN+"))


ggplot(df, aes(x = year, y = cohort, group = cohort)) + 
   # geom_mark_rect(data = cs_df, aes(group = year), fill = "#00BFC4", color = NA, expand = unit(7, "mm"), ) +
    geom_mark_rect(data = wave_df, aes(group = group, fill = phdcn), color = NA, expand = unit(6, "mm")) +
  geom_line(size = 3, aes(color = focus)) +
  geom_richtext(aes(label = age, fill = focus), 
                size = 5, label.colour = NA, text.color = "black") +
  theme_minimal(base_size = 24) +
  scale_fill_manual(values = c("yes" = "white", "no" = "grey50", "PHDCN" = "#F8766D", "PHDCN+" = "#00BFC4")) +
  scale_color_manual(values = c("yes" = "white", "no" = "grey50", "PHDCN" = "#F8766D", "PHDCN+" = "#00BFC4")) +
  labs(y = "Cohort Ages", x= NULL) +
  scale_x_continuous(breaks = seq(1995, 2021, by = 5), limits = c(1994,2022)) +
  scale_y_continuous(limits = c(-1,22)) +
  
  annotate("text", x = 1998.5, y = 21.25, family = plot_font, label = "Original PHDCN\nWaves 1–3", color = "#F8766D", size = 5) +
  annotate("text", x = 2016.5, y = 21.25, family = plot_font, label = "PHDCN+\nWaves 4 & 5", color = "#00BFC4", size = 5) +
  theme(panel.grid = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(color = "grey90"),
        text = element_text(family = plot_font,  color = "white"),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        legend.position = "none")
```

::: {.notes}
PHDCN+ covers much longer time period than original PHDCN: Age 0 to 40, 25 years within each cohort

Well past peak criminal involvement and exposure ages

Staggered cohort allows separation of age and period effects---we'll cover this next session
:::

## Contexts of Violence

```{r}
#| fig.width: 9
#| fig.height: 6
load("./data/cohort_age_hom.RData")
axis_data <- tibble(cohort = factor(c(0, 9, 12, 15), levels = c(0, 9, 12, 15)),
                          hom_rate = rep(24.8,4),
                          age = rep(-3, 4))
cohort_age_hom %>%
  mutate(cohort = str_remove(cohort, "Cohort "),
         cohort = fct_reorder(cohort, as.numeric(cohort)),
         cohort = fct_rev(cohort)) %>%
  ggplot(aes(x = age, y = hom_rate, color = cohort)) + 
  geom_line() + 
  facet_wrap(~cohort, strip.position = "left", ncol = 1) +  
  scale_y_continuous(breaks = c(20, 30), 
                     position = "right", 
                     sec.axis = sec_axis(trans = ~.*0, name = "Cohort")) +
  annotate("rect", xmin = 9, xmax = 25, ymin = -Inf, ymax = Inf, fill = "white", alpha = 0.3) +
  theme_minimal(base_size = 24) +
  geom_text(data = axis_data,
            aes(label = cohort), size = 7) +
  coord_cartesian(clip = FALSE, xlim = c(0,40)) +
  labs(x = "Cohort member age", y = "Homicide Rate") +
  theme(panel.grid = element_blank(),
        axis.text = element_text(color = "grey90"),
        axis.text.y.right =  element_blank(),
        axis.text.y.left  = element_blank(),
        strip.text.y.left = element_text(color =  "transparent", angle = 0),
        legend.position = "none",
        text = element_text(family = plot_font,  color = "white"),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA))
```

::: {.notes}
Align by age instead of year---see age-specific violence context

Age 9 to 25 highlighted as peak risk period

4 different patterns: High decline, medium decline, stable low, rising

Note that many other changes were occurring than just violence---incarceration, arrests, the economy. I'll discuss those more in the talk in the next session.
:::

# Life-Course Exposures to Gun Violence {background-image="img/ozzie-stern-dkwJLowVvl4-unsplash.jpg" background-opacity="0.3"}

## Measures {background-image="img/ozzie-stern-dkwJLowVvl4-unsplash.jpg" background-opacity="0.1"}

-   Has the respondent ever seen someone else get shot? 
    -   If so, at what age? (only wave 2)
-   Has the respondent ever been shot?
    -   If so, at what age? (waves 2 & 5)

. . .

Problem: Interval censoring

```{r}
#| fig.height: 0.9
#| fig.width: 8

tibble(
  Age = c(0, 3, 6, 17, 17, 26),
  red = c("white", "white", "white", "white", "red", "red"),
  `Seen Shot` = c("No", "No", "No", "No", "No", "Yes"),
  y = c(1,1,1,1,1, 1)
) %>%
  ggplot(aes(x = Age, y = y)) + 
  geom_line(aes(color = red), size = 2) + 
  annotate("text", x = -6, y = 1, label = "Seen Shot", color = "white", size = 7) +
  annotate("text", x = -6, y = 0, label = "Age", color = "white", size = 7) +
  geom_text(aes(x = Age, label = Age, y = y - 1), color = "white", size = 7) +
  scale_color_manual(values = c("red" = "red", "white" = "white")) +
  geom_richtext(aes(label = `Seen Shot`), size = 7, label.colour = NA, text.color = "black") +
  coord_cartesian(ylim =  c(-0.7, 1.7), xlim = c(-9, 28), expand = FALSE) +
  theme_void(base_size = 24) +
  theme(legend.position = "none",
        text = element_text(family = plot_font,  color = "white"),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA))
```

. . .

Method: Non-parametric MLE (Turnbull 1976)


::: {.notes}
Interval censoring means we don't know exactly when they saw someone shot if they

Data also right censored for people who dropped out of study

Turnbull method makes no statistical assumptions, just calculates cumulative probability of exposure by age accounting for censoring.
:::

## Seen shot: race/ethnicity

```{r}
#| fig.width: 9
#| fig.height: 6
gg_cols <- gg_color_hue(3)
par(family = plot_font, bg = NA, cex = 1.75, mar = c(4,4,1,0))
plot(npt_race_models[["seen_shot"]], dtype = "cdf", 
     COL = FALSE, 
     XLAB = "", 
     YLAB = "",
     LTY = "solid",
     ylim = c(0, 0.75),
     bty="n",
     LEGEND = FALSE,
     axes = FALSE,
     main = NULL,
     estpar = list(col = gg_cols, lwd = 1.5))
axis(1, col = NA, tick = FALSE, col.axis = "white")
axis(2, col = NA, tick = FALSE, las = 1, col.axis = "white", at = seq(0,0.75, by= 0.25))
# grid(nx = NULL, ny = NA, lty = "dashed", col = "grey90", lwd = 1)
title(xlab = "Age", ylab = "Cumulative Probability",  col.lab = "white")
text(x = rep(0, 5),
     adj = 0,
     font = c(1, 1, 1 ,2),
     y = c(seq(0.53, 0.68, length.out = 3), 0.75),
     labels = c("NH White", "NH Black", "Hispanic", "Race/Ethnicity"),
     col = c(gg_cols, "white"))
```

::: {.notes}
Long plateaus are due to interval censoring: The upticks immediately afterward occurred somewhere in that period.

Note that seeing people shot typically occurs before age 20

60% of black and hispanic respondents seen someone shot by their 40s. Under 30% for whites.

While exposure lower for white respondents, it still ticks up in later ages, which are all the recent period of elevated violence
:::

## Been shot: race/ethnicity

```{r}
#| fig.width: 9
#| fig.height: 6
par(family = plot_font, bg = NA, cex = 1.75, mar = c(4,4,1,0))
plot(npt_race_models[["been_shot"]], dtype = "cdf", 
     COL = FALSE, 
     XLAB = "", 
     YLAB = "",
     LTY = "solid",
     ylim = c(0, 0.09),
     bty="n",
     LEGEND = FALSE,
     axes = FALSE,
     main = NULL,
     estpar = list(col = gg_cols, lwd = 1.5))
axis(1, col = NA, tick = FALSE, col.axis = "white")
axis(2, col = NA, tick = FALSE, las = 1, col.axis = "white", at = seq(0, 0.09, by= 0.03))
# grid(nx = NULL, ny = NA, lty = "dashed", col = "grey90", lwd = 1)
title(xlab = "Age", ylab = "Cumulative Probability",  col.lab = "white")
text(x = rep(0, 5),
     adj = 0,
     font = c(1, 1, 1 ,2),
     y = c(seq(0.055, 0.082, length.out = 4)[-1], 0.09),
     labels = c("NH White", "NH Black", "Hispanic", "Race/Ethnicity"),
     col = c(gg_cols, "white"))
```

::: {.notes}
Being shot occurs later on average---mostly after 18 or so

No white respondents were shot after age 22, while black and Hispanic respondents saw exposures in recent period
:::



## Seen shot: sex

```{r}
#| fig.width: 9
#| fig.height: 6
par(family = plot_font, bg = NA, cex = 1.75, mar = c(4,4,1,0))
gg_cols <- gg_color_hue(2)
plot(npt_sex_models[["seen_shot"]], dtype = "cdf", 
     COL = FALSE, 
     XLAB = "", 
     YLAB = "",
     LTY = "solid",
     ylim = c(0, 0.6),
     bty="n",
     LEGEND = FALSE,
     axes = FALSE,
     main = NULL,
     estpar = list(col = gg_cols, lwd = 1.5))
axis(1, col = NA, tick = FALSE, col.axis = "white")
axis(2, col = NA, tick = FALSE, las = 1, col.axis = "white", at = seq(0, 0.6, by= 0.2))
# grid(nx = NULL, ny = NA, lty = "dashed", col = "grey90", lwd = 1)
title(xlab = "Age", ylab = "Cumulative Probability",  col.lab = "white")
text(x = rep(0, 3),
     adj = 0,
     font = c(1, 1, 2),
     y = c(seq(0.5, 0.55, length.out = 2), 0.6),
     labels = c("Female", "Male", "Sex"),
     col = c(gg_cols, "white"))
```

::: {.notes}
Men are somewhat more likely to see a shooting, but general pattern is similar across age

Again, interval censoring plateaus
:::


## Been shot: sex

```{r}
#| fig.width: 9
#| fig.height: 6
par(family = plot_font, bg = NA, cex = 1.75, mar = c(4,4,1,0))
gg_cols <- gg_color_hue(2)
plot(npt_sex_models[["been_shot"]], dtype = "cdf", 
     COL = FALSE, 
     XLAB = "", 
     YLAB = "",
     LTY = "solid",
     ylim = c(0, 0.12),
     bty="n",
     LEGEND = FALSE,
     axes = FALSE,
     main = NULL,
     estpar = list(col = gg_cols, lwd = 1.5))
axis(1, col = NA, tick = FALSE, col.axis = "white")
axis(2, col = NA, tick = FALSE, las = 1, col.axis = "white", at = seq(0, 0.12, by= 0.04))
# grid(nx = NULL, ny = NA, lty = "dashed", col = "grey90", lwd = 1)
title(xlab = "Age", ylab = "Cumulative Probability",  col.lab = "white")
text(x = rep(0, 3),
     adj = 0,
     font = c(1, 1, 2),
     y = c(seq(0.1, 0.11, length.out = 2), 0.12),
     labels = c("Female", "Male", "Sex"),
     col = c(gg_cols, "white"))
```

::: {.notes}
Sex divergence is very large for being shot

Basically flat for women after teens, though a jump in mid-30s, possibly DV? We're investigating
:::

# Risk and Protective Factors {background-image="img/clay-banks-nsAn3nSW5T0-unsplash.jpg" background-opacity="0.3"}

## Additional Measures {background-image="img/clay-banks-nsAn3nSW5T0-unsplash.jpg" background-opacity="0.1"}

::: {.nonincremental}
- Cohort
- Immigrant generation
- Childhood neighbourhood context
    -   Collective efficacy
    -   Disadvantage
    -   Homicide rate
:::

. . .

Method: Semi-parametric proportional hazards MLE (Anderson-Bergman 2017)

## Hazard Model Estimates

```{r, icsp-models-estimates}
rename_terms_and_outcome <- function(x){
   out_df <- mutate(x, term = 
             case_when(
               str_detect(term, "race") ~ str_replace(term, "race", "Race "),
               str_detect(term, "sex") ~ "Male",
               str_detect(term, "cohort_0Older Cohorts:whiteWhite") ~ "Older Cohorts x White",
               str_detect(term, "cohort_0Older Cohorts") ~ "Older Cohorts",
               str_detect(term, "white") ~ str_remove(term, "white"),
               str_detect(term, "cohort") ~ str_c("Cohort ", str_remove(term, "cohort")),
               str_detect(term, "pcimgen") ~ str_replace(term, "pcimgen", "Immig. Gen "),
               str_detect(term, "PCA_disadv") ~ str_c("Disadvantage (Age ", str_extract(term, "[1-9]*$"), ")"),
               str_detect(term, "hom_rate") ~ str_c("Homicide Rate (Age ", str_extract(term, "[1-9]*$"), ")"),
               str_detect(term, "collective_efficacy") ~ str_c("Collective Efficacy (Age ", str_extract(term, "[1-9]*$"), ")"),
               TRUE ~ "ERROR"
           ),
         outcome = str_to_title(str_replace(outcome, "_"," ")))
   return(out_df)
}

map_dfr(c("seen_shot", "been_shot"), ~ get_spt_estimates(spt_base_models, .x, zval = 1.96)) %>%
   rename_terms_and_outcome() %>%
   mutate(term = fct_rev(fct_relevel(term, 
                              "Race Black", "Race Hispanic", "Race Other", 
                              "Male", 
                              "Immig. Gen 2", "Immig. Gen 1",
                              "Homicide Rate (Age 9)",
                              "Disadvantage (Age 9)",
                              "Collective Efficacy (Age 9)",
                              "Cohort 12", "Cohort 9", "Cohort 0",))) %>%
  ggplot(aes(x = estimate, y = term)) + 
    facet_wrap(~outcome, nrow = 1) +
    geom_vline(xintercept=1, linetype = "dashed", alpha = 0.8, size = 0.5, color = "grey80") +
    geom_point(size =1, color = "white") + 
    scale_x_log10(breaks = c(0.1, 1, 10), labels = c("0.1", "1.0", "10")) +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), size =0.25, height = 0.25, color = "white") +
    labs(y = NULL, x = "Hazard Ratio", title = NULL, subtitle = NULL) +
    theme_minimal(base_size = 18) + 
    coord_cartesian(xlim = c(0.1, 10)) +
    theme(panel.grid = element_blank(),
        axis.text = element_text(color = "grey90", family = plot_font),
        strip.text = element_text(color =  "white", family = plot_font, size = rel(1.2)),
        legend.position = "none",
        text = element_text(family = plot_font,  color = "white"),
        panel.spacing.x = unit(0.8, "in"),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA))
```

::: notes
Hazard ratios are multiplicative risk changes; x more likely to experience event at any given moment

Been shot is rare, so statistical power is low

Most patterns similar across outcomes, except cohort 0

Here the cohorts are proxying for any historical differences

This includes different overall city levels of violence

Very modest relationship between exposure and neighborhood measures; trumped by individual factors
:::

## Predictions: Individuals

```{r}
#| fig.width: 9
#| fig.height: 4.5
logit_models <- list(
`Seen Shot` = glm(formula = factor(seen_shot_ever) ~ cohort + sex + race + 
    PCA_disadv_09 + hom_rate_avg_09 + collective_efficacy_09 + pcimgen, family = binomial, data = survival_data[["seen_shot"]] %>% filter(w5_sample)),
`Been Shot` = glm(formula = factor(been_shot_ever) ~ cohort + sex + race + 
    PCA_disadv_09 + hom_rate_avg_09 + collective_efficacy_09 + pcimgen, family = binomial, data = survival_data[["been_shot"]] %>% filter(w5_sample)))

map_dfr(logit_models, ~tidy(marginalmeans(., type = "response")), .id = "outcome") %>%
  filter(term == "race" | term == "sex") %>%
  filter(value != "Other") %>%
  mutate(
    value = fct_recode(value, Male = "m", Female = "f"),
    term = str_to_title(term)) %>%
  ggplot(aes(x = value, y = estimate, fill = value)) + 
  geom_col() + 
  facet_grid(outcome~ term, scales = "free") +
  labs(x = NULL, y = "Probability", title = NULL) +
  scale_y_continuous(breaks = scales::breaks_pretty()) +
  theme_minimal(base_size = 24) +
  theme(panel.grid = element_blank(),
           axis.text = element_text(family = plot_font,  color = "white"),
          # axis.text.y.right =  element_blank(),
          # axis.text.y.left  = element_blank(),
          strip.background = element_rect(fill = "transparent",colour = NA),
          strip.text = element_text(family = plot_font,  color = "white"),
          legend.position = "none",
          text = element_text(family = plot_font,  color = "white"),
          panel.background = element_rect(fill = "transparent",colour = NA),
          plot.background = element_rect(fill = "transparent",colour = NA))
```
::: aside
*Predicted probability of ever been or seen shot, other variables at mean*
:::

## Predictions: Neighborhoods

```{r}
#| fig.width: 9
#| fig.height: 4.5
map_dfr(logit_models, ~tidy(marginaleffects(., type = "response", newdata = "mean")), .id = "outcome") %>%
  filter(str_detect(term, "^(PCA|hom|coll)")) %>%
  mutate(term = case_when(
    str_detect(term, "PCA") ~ "Disadvantage",
    str_detect(term, "hom") ~ "Homicide",
    str_detect(term, "coll") ~ "Collective Efficacy",
    TRUE ~ "ERROR"
  )) %>%
  mutate(term = factor(term, levels = c("Disadvantage", "Collective Efficacy", "Homicide"))) %>%
  ggplot(aes(x = term, y = estimate, fill = term)) + 
  geom_col() + 
  facet_wrap(~outcome, ncol = 2, scales = "fixed") +
  labs(x = NULL, y = "Probability", title = NULL) +
  scale_y_continuous(breaks = scales::breaks_pretty()) +
    scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
  theme_minimal(base_size = 24) +
  theme(panel.grid = element_blank(),
          axis.text = element_text(family = plot_font,  color = "white"),
          # axis.text.y.right =  element_blank(),
          # axis.text.y.left  = element_blank(),
          strip.background = element_rect(fill = "transparent",colour = NA),
          strip.text = element_text(family = plot_font,  color = "white"),
          legend.position = "none",
          text = element_text(family = plot_font,  color = "white"),
          panel.background = element_rect(fill = "transparent", colour = NA),
          plot.background = element_rect(fill = "transparent",colour = NA))
```

::: aside
*Predicted marginal effect on ever been or sheet shot, other variables at means*
:::


## Takeaways {background-image="img/joel-mott-s-rsM-AktbA-unsplash.jpg" background-opacity="0.1" .smaller}

- Being shot occurs later in life than seeing shootings

- Only Black and Hispanic respondents were shot in adulthood

- Seeing shootings much less common for white respondents

- Mainly males are shot, but female respondents witness nearly as many shootings

- Neighbourhood context during childhood consequential mainly for seeing shootings

## Feedback and Questions {background-image="img/joel-mott-s-rsM-AktbA-unsplash.jpg" background-opacity="0.1" .smaller}

Contact:

| Charles C. Lanfear
| Institute of Criminology
| University of Cambridge
| [cl948@cam.ac.uk](mailto:cl948@cam.ac.uk)

For more about the PHDCN+:

| PHDCN@fas.harvard.edu
| [https://sites.harvard.edu/phdcn/](https://sites.harvard.edu/phdcn/)
| [https://doi.org/10.1007/s40865-022-00203-0](https://sites.harvard.edu/phdcn/)


# Appendix {background-image="img/ben-sp-fD4AwrWv4t8-unsplash-crop.jpg" background-opacity="0.3"}

## References

Sampson, Kirk, & Bucci. 2022. "Cohort Profile: Project on Human Development in Chicago Neighborhoods and Its Additions (PHDCN+)." *Journal of Developmental and Life-Course Criminology* 8.

Turnbull. 1976. "The empirical distribution function with arbitrarily grouped, censored and truncated data." *Journal of the Royal Statistical Society: Series B (Methodological)* 38(3):290-295.

Anderson-Bergman. 2017. "icenReg: Regression models for interval censored data in R." *Journal of Statistical Software* 81(12):1--23.
