---
format:
  revealjs:
    theme: night
    logo: img/phdcn_logo_white.svg
    incremental: true
editor: visual
mouse-wheel: true
history: false
---

```{r setup}
#| include: false
library(tidyverse)
library(ggforce)
library(ggtext)
library(showtext)
library(icenReg)
library(interval)
load("./data/survival_data.RData")
load("../../../Kirk Projects/phdcn_gun_violence_exposures/data/output/npt_cohort_models.RData")
load("../../../Kirk Projects/phdcn_gun_violence_exposures/data/output/npt_race_models.RData")
load("../../../Kirk Projects/phdcn_gun_violence_exposures/data/output/spt_base_models.RData")
source("../../../Kirk Projects/phdcn_gun_violence_exposures/syntax/project_functions.R")
knitr::opts_chunk$set(dev = "svg",  
                      dev.args = list(bg = 'transparent'),
                      message = FALSE,
                      echo = FALSE,
                      warning = FALSE,
                      fig.showtext = TRUE,
                      fig.retina = 1,
                      fig.align = "center")
plot_font <- "Open Sans"
get_age_percents <- function(x, .race = NULL, .cohort = NULL){
  if(is.null(.race) & is.null(.cohort)){
    exposure_ecdf <- ecdf(x$age_right)
    return(exposure_ecdf(1:40))
  } else if (is.character(.race)){
    exposure_ecdf <- ecdf(x$age_right[x$race == .race])
    return(exposure_ecdf(1:40))
  } else if (is.character(.cohort)){
    exposure_ecdf <- ecdf(x$age_right[x$cohort == .cohort])
    return(exposure_ecdf(1:40))
  } else {
    stop("If provided, race or cohort must be a character value")
  }
}
```

```{r}
#| include: false
#| cache: false

font_add_google(name = plot_font)

```

## The Changing Risks of Exposure to Gun Violence in Chicago {background-image="img/stephan-cassara-KnAIsBuitGg-unsplash.jpg" background-opacity=0.3}

Charles C. Lanfear

Robert J. Sampson

David S. Kirk

Rebecca Bucci


## Homicide Rates in Chicago

```{r}
#| fig.width: 9
#| fig.height: 6
load("./data/hom_rate_df.RData")
hom_rate_df %>%
  select(year, gun_hom_rate, hom_rate) %>%
  pivot_longer(-year) %>%
  mutate(name = ifelse(str_detect(name, "gun"), "Gun\nHomicide", "All\nHomicide")) %>%
  filter(!is.na(value)) %>%
  ggplot(aes(x = year, y = value, group = name, color = name)) + 
  geom_line(size = 1) +
  labs(x = NULL, y = "Homicides per 100,000") +
  scale_y_continuous(limits = c(0, NA)) +
  geom_text(data = tibble(name = c("Gun\nHomicide", "All\nHomicide"),
                          year = c(1985, 1985),
                          value = c(10, 31)),
            aes(label = name),
            size = 6, family = plot_font) +
  theme_minimal(base_size = 24) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.text.x = element_text(color = "grey90"),
        axis.text.y = element_text(color = "grey90"),
        axis.title.y = element_text(color = "grey90"),
        axis.ticks.y = element_line(color = "grey90"), 
        text = element_text(family = plot_font,  color = "white"),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        legend.position = "none")
```

::: {.notes}
Enormous variation in homicide; early 1990s peak was double the period 2005-2015

Resurgence since 2017

Even during low periods, this is ten or more times the homicide rate of major European cities 
:::


## Questions {background-image="img/stephan-cassara-KnAIsBuitGg-unsplash.jpg" background-opacity=0.1}

&nbsp;

::: {.fragment}
How were children growing up in this period exposed to **gun violence**?
:::


&nbsp;

::: {.fragment}
And how does this exposure differ... 
:::

::: {.fragment}
| ... by **cohort**?
:::

::: {.fragment}
| ... by **race** and **sex**?
:::

::: {.fragment}
| ... by **neighbourhood** context?
:::


::: {.notes}
We'll be focusing mostly on cohort and race here.
:::

# The {background-image="img/max-bender-yb7Yg3Rv7WA-unsplash.jpg" background-opacity=0.3}

![](img/phdcn_logo_white.svg)


## {background-image="img/max-bender-yb7Yg3Rv7WA-unsplash.jpg" background-opacity=0.1}

### Project on Human Development in Chicago Neighborhoods

* 6200 children in 7 birth cohorts
* Interviewed in 3 waves from 1995–2002
* Community surveys in 1995 and 2002

::: {.fragment}
### PHDCN+
:::

* Representative sample of 4 cohorts
* 1057 interviewed in 2012 
* 682 followed-up in 2021
* Focus: **Gun Violence**

::: {.footer}
See Sampson, Kirk, & Bucci (2022)
:::

## Timeline

```{r}
#| fig.width: 9
#| fig.height: 6
df <- tribble(
  ~cohort, ~time, ~year, ~age, 
  0,  1, 1995,  0,
  0,  2, 2021, 25,
  3,  1, 1995,  3,
  3,  2, 2002, 11,
  6,  1, 1995,  6,
  6,  2, 2002, 14,
  9,  1, 1995, 9, 
  9,  2, 2021, 34,
  12, 1, 1995, 12,
  12, 2, 2021, 37,
  15, 1, 1995, 15,
  15, 2, 2021, 40,
  18, 1, 1995, 18,
  18, 2, 2002, 26,
) %>%
  mutate(focus = ifelse(cohort %in% c(0,9,12,15), "yes", "no"))

cs_df <- tibble(
  survey = rep(c("PHDCN-CS", "CCAHS"), each = 7),
  year   = rep(c(1995, 2002), each = 7),
  cohort = rep(seq(0,18, by = 3), length.out = 14)
)
wave_df <- tibble(
  survey = rep(1:5, length.out = 35),
  year   = rep(c(1995, 1998.5, 2002, 2012, 2021), length.out = 35),
  cohort = rep(seq(0,18, by = 3), each = 5)
) %>%
  filter(year <= 2002 | cohort %in% c(0, 9, 15)) %>%
  mutate(group = 
           case_when(
            year <= 2002 ~ year, 
            year > 2002 & cohort %in% c(9,15,18) ~ year,
            year > 2002 & cohort == 0 ~ year -1),
         phdcn = ifelse(year <= 2002, "PHDCN", "PHDCN+"))


ggplot(df, aes(x = year, y = cohort, group = cohort)) + 
   # geom_mark_rect(data = cs_df, aes(group = year), fill = "#00BFC4", color = NA, expand = unit(7, "mm"), ) +
    geom_mark_rect(data = wave_df, aes(group = group, fill = phdcn), color = NA, expand = unit(6, "mm")) +
  geom_line(size = 3, aes(color = focus)) +
  geom_richtext(aes(label = age, fill = focus), 
                size = 5, label.colour = NA, text.color = "black") +
  theme_minimal(base_size = 24) +
  scale_fill_manual(values = c("yes" = "white", "no" = "grey50", "PHDCN" = "#F8766D", "PHDCN+" = "#00BFC4")) +
  scale_color_manual(values = c("yes" = "white", "no" = "grey50", "PHDCN" = "#F8766D", "PHDCN+" = "#00BFC4")) +
  labs(y = "Cohort Ages", x= NULL) +
  scale_x_continuous(breaks = seq(1995, 2021, by = 5), limits = c(1994,2022)) +
  scale_y_continuous(limits = c(-1,22)) +
  
  annotate("text", x = 1998.5, y = 21.25, family = plot_font, label = "Original PHDCN\nWaves 1–3", color = "#F8766D", size = 5) +
  annotate("text", x = 2016.5, y = 21.25, family = plot_font, label = "PHDCN+\nWaves 4 & 5", color = "#00BFC4", size = 5) +
  theme(panel.grid = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(color = "grey90"),
        text = element_text(family = plot_font,  color = "white"),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        legend.position = "none")
```

::: {.notes}
PHDCN+ covers much longer time period than original PHDCN: Age 0 to 40, 25 years within each cohort

Well past peak criminal involvement and exposure ages
:::

## Contexts of Violence

```{r}
#| fig.width: 9
#| fig.height: 6
load("./data/cohort_age_hom.RData")
axis_data <- tibble(cohort = factor(c(0, 9, 12, 15), levels = c(0, 9, 12, 15)),
                          hom_rate = rep(24.8,4),
                          age = rep(-3, 4))
cohort_age_hom %>%
  mutate(cohort = str_remove(cohort, "Cohort "),
         cohort = fct_reorder(cohort, as.numeric(cohort)),
         cohort = fct_rev(cohort)) %>%
  ggplot(aes(x = age, y = hom_rate, color = cohort)) + 
  geom_line() + 
  facet_wrap(~cohort, strip.position = "left", ncol = 1) +  
  scale_y_continuous(breaks = c(20, 30), 
                     position = "right", 
                     sec.axis = sec_axis(trans = ~.*0, name = "Cohort")) +
  annotate("rect", xmin = 13, xmax = 25, ymin = -Inf, ymax = Inf, fill = "white", alpha = 0.3) +
  theme_minimal(base_size = 24) +
  geom_text(data = axis_data,
            aes(label = cohort), size = 7) +
  coord_cartesian(clip = FALSE, xlim = c(0,40)) +
  labs(x = "Cohort member age", y = "Homicide Rate") +
  theme(panel.grid = element_blank(),
        axis.text = element_text(color = "grey90"),
        axis.text.y.right =  element_blank(),
        axis.text.y.left  = element_blank(),
        strip.text.y.left = element_text(color =  "transparent", angle = 0),
        legend.position = "none",
        text = element_text(family = plot_font,  color = "white"),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA))
```


::: {.notes}
Align by age instead of year---see age-specific violence context

Age 13 to 25 highlighted as peak risk period

4 different patterns: High decline, medium decline, stable low, rising
:::


# Lifecourse Exposures to Gun Violence {background-image="img/ozzie-stern-dkwJLowVvl4-unsplash.jpg" background-opacity=0.3}


## Measures {background-image="img/ozzie-stern-dkwJLowVvl4-unsplash.jpg" background-opacity=0.1}

Has the respondent...

* ever been shot?
* ever seen someone else get shot?


. . .

&nbsp;

Problem: Interval censoring

. . .

&nbsp;

Method: Non-parametric MLE (Turnbull 1976)

## Been shot: cohort

```{r}
#| fig.width: 9
#| fig.height: 6
par(family = plot_font, bg = NA, cex = 1.75, mar = c(4,4,1,0))
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

gg_cols <- gg_color_hue(4)
plot(npt_cohort_models[["been_shot"]], dtype = "cdf", 
     COL = FALSE, 
     XLAB = "", 
     YLAB = "",
     LTY = "solid",
     ylim = c(0, 0.09),
     bty="n",
     LEGEND = FALSE,
     axes = FALSE,
     main = NULL,
     estpar = list(col = gg_cols, lwd = 1.5))
axis(1, col = NA, tick = FALSE, col.axis = "white")
axis(2, col = NA, tick = FALSE, las = 1, col.axis = "white", at = seq(0, 0.09, by= 0.03))
# grid(nx = NULL, ny = NA, lty = "dashed", col = "grey90", lwd = 1)
title(xlab = "Age", ylab = "Cumulative Probability",  col.lab = "white")
text(x = rep(0, 5),
     adj = 0,
     font = c(2, 1, 1 ,1, 1),
     y = rev(c(seq(0.055, 0.082, length.out = 4), 0.09)),
     labels = c("Cohort", 15, 12, 9, 0),
     col = c("white", gg_cols))
```

::: {.notes}
Being shot happens mainly in young adulthood: 17 to 25

Cohort 0 has unusually high early age likelihood of being shot, but looks similar to 12 and 15 later on.

:::



## Seen shot: cohort

```{r}
#| fig.width: 9
#| fig.height: 6
par(family = plot_font, bg = NA, cex = 1.75, mar = c(4,4,1,0))
plot(npt_cohort_models[["seen_shot"]], dtype = "cdf", 
     COL = FALSE, 
     XLAB = "", 
     YLAB = "",
     LTY = "solid",
     ylim = c(0, 0.9),
     bty="n",
     LEGEND = FALSE,
     axes = FALSE,
     main = NULL,
     estpar = list(col = gg_cols, lwd = 1.5))
axis(1, col = NA, tick = FALSE, col.axis = "white")
axis(2, col = NA, tick = FALSE, las = 1, col.axis = "white", at = seq(0,0.9, by= 0.3))
# grid(nx = NULL, ny = NA, lty = "dashed", col = "grey90", lwd = 1)
title(xlab = "Age", ylab = "Cumulative Probability",  col.lab = "white")
text(x = rep(0, 5),
     adj = 0,
     font = c(2, 1, 1 ,1, 1),
     y = rev(c(seq(0.55, 0.82, length.out = 4), 0.9)),
     labels = c("Cohort", 15, 12, 9, 0),
     col = c("white", gg_cols))
```

::: {.notes}
In contrast to being shot, witnessing shootings occurs mainly before age 20

Due to interval censoring, focus mainly on 9, 12, 15. Fairly similar. Hard to tell here, but cohort zero is unusual with low exposure.
:::


## Been shot: race/ethnicity

```{r}
#| fig.width: 9
#| fig.height: 6
par(family = plot_font, bg = NA, cex = 1.75, mar = c(4,4,1,0))
gg_cols <- gg_color_hue(3)
plot(npt_race_models[["been_shot"]], dtype = "cdf", 
     COL = FALSE, 
     XLAB = "", 
     YLAB = "",
     LTY = "solid",
     ylim = c(0, 0.09),
     bty="n",
     LEGEND = FALSE,
     axes = FALSE,
     main = NULL,
     estpar = list(col = gg_cols, lwd = 1.5))
axis(1, col = NA, tick = FALSE, col.axis = "white")
axis(2, col = NA, tick = FALSE, las = 1, col.axis = "white", at = seq(0, 0.09, by= 0.03))
# grid(nx = NULL, ny = NA, lty = "dashed", col = "grey90", lwd = 1)
title(xlab = "Age", ylab = "Cumulative Probability",  col.lab = "white")
text(x = rep(0, 5),
     adj = 0,
     font = c(1, 1, 1 ,2),
     y = c(seq(0.055, 0.082, length.out = 4)[-1], 0.09),
     labels = c("NH White", "NH Black", "Hispanic", "Race/Ethnicity"),
     col = c(gg_cols, "white"))
```

::: {.notes}
Illustrates again that being shot occurs later

No white respondents were shot after age 22, while black and Hispanic respondents saw exposures in recent period
:::

## Seen shot: race/ethnicity

```{r}
#| fig.width: 9
#| fig.height: 6
par(family = plot_font, bg = NA, cex = 1.75, mar = c(4,4,1,0))
plot(npt_race_models[["seen_shot"]], dtype = "cdf", 
     COL = FALSE, 
     XLAB = "", 
     YLAB = "",
     LTY = "solid",
     ylim = c(0, 0.9),
     bty="n",
     LEGEND = FALSE,
     axes = FALSE,
     main = NULL,
     estpar = list(col = gg_cols, lwd = 1.5))
axis(1, col = NA, tick = FALSE, col.axis = "white")
axis(2, col = NA, tick = FALSE, las = 1, col.axis = "white", at = seq(0,0.9, by= 0.3))
# grid(nx = NULL, ny = NA, lty = "dashed", col = "grey90", lwd = 1)
title(xlab = "Age", ylab = "Cumulative Probability",  col.lab = "white")
text(x = rep(0, 5),
     adj = 0,
     font = c(1, 1, 1 ,2),
     y = c(seq(0.55, 0.82, length.out = 4)[-1], 0.9),
     labels = c("NH White", "NH Black", "Hispanic", "Race/Ethnicity"),
     col = c(gg_cols, "white"))
```


::: {.notes}
Long plateaus are due to interval censoring: The upticks immediately afterward occurred somewhere in that period.

Again, we see Seeing someone shot happens earlier than being shot

Overall exposure lower for white respondents but, unlike being shot, ticks up in later ages, which are all the recent period
:::


# Risk and Protective Factors {background-image="img/clay-banks-nsAn3nSW5T0-unsplash.jpg" background-opacity=0.3}

## Additional Measures {background-image="img/clay-banks-nsAn3nSW5T0-unsplash.jpg" background-opacity=0.1}

* Sex
* Immigrant generation
* Childhood neighbourhood context
   * Collective efficacy
   * Disadvantage
   * Homicide rate

. . .

Method: Semi-parametric proportional hazards MLE (Anderson-Bergman 2017)

## Estimates

```{r, icsp-models-estimates}
rename_terms_and_outcome <- function(x){
   out_df <- mutate(x, term = 
             case_when(
               str_detect(term, "race") ~ str_replace(term, "race", "Race "),
               str_detect(term, "sex") ~ "Male",
               str_detect(term, "cohort_0Older Cohorts:whiteWhite") ~ "Older Cohorts x White",
               str_detect(term, "cohort_0Older Cohorts") ~ "Older Cohorts",
               str_detect(term, "white") ~ str_remove(term, "white"),
               str_detect(term, "cohort") ~ str_c("Cohort ", str_remove(term, "cohort")),
               str_detect(term, "pcimgen") ~ str_replace(term, "pcimgen", "Immig. Gen "),
               str_detect(term, "PCA_disadv") ~ str_c("Disadvantage (Age ", str_extract(term, "[1-9]*$"), ")"),
               str_detect(term, "hom_rate") ~ str_c("Homicide Rate (Age ", str_extract(term, "[1-9]*$"), ")"),
               str_detect(term, "collective_efficacy") ~ str_c("Collective Efficacy (Age ", str_extract(term, "[1-9]*$"), ")"),
               TRUE ~ "ERROR"
           ),
         outcome = str_to_title(str_replace(outcome, "_"," ")))
   return(out_df)
}

map_dfr(c("seen_shot", "been_shot"), ~ get_spt_estimates(spt_base_models, .x, zval = 1.96)) %>%
   rename_terms_and_outcome() %>%
   mutate(term = fct_rev(fct_relevel(term, 
                              "Cohort 12", "Cohort 9", "Cohort 0",
                              "Male", 
                              "Race Black", "Race Hispanic", "Race Other", 
                              "Immig. Gen 2", "Immig. Gen 1"))) %>%
  ggplot(aes(x = estimate, y = term)) + 
    facet_wrap(~outcome, nrow = 1) +
    geom_vline(xintercept=1, linetype = "dashed", alpha = 0.8, size = 0.5, color = "grey80") +
    geom_point(size =1, color = "white") + 
    scale_x_log10(breaks = c(0.1, 1, 10), labels = c("0.1", "1.0", "10")) +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), size =0.25, height = 0.25, color = "white") +
    labs(y = NULL, x = "Hazard Ratio", title = NULL, subtitle = NULL) +
    theme_minimal(base_size = 18) + 
    coord_cartesian(xlim = c(0.1, 10)) +
    theme(panel.grid = element_blank(),
        axis.text = element_text(color = "grey90", family = plot_font),
        strip.text = element_text(color =  "white", family = plot_font, size = rel(1.2)),
        legend.position = "none",
        text = element_text(family = plot_font,  color = "white"),
        panel.spacing.x = unit(0.8, "in"),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA))
```

::: {.notes}

Been shot is rare, so statistical power is low

Most patterns similar across outcomes, except cohort 0

Here the cohorts are proxying for any historical differences

This includes different overall city levels of violence

Very modest relationship between exposure and neighborhood measures; trumped by individual factors

:::

# Takeaways {background-image="img/joel-mott-s-rsM-AktbA-unsplash.jpg" background-opacity=0.1}

* Being shot tends to occur later in life than seeing shootings

  * Seeing shootings more likely when homicide is high in adolescence
  * Being shot more likely when homicide is high in early adulthood

* Risk of being shot continues throughout adulthood for black and Hispanic but not white respondents

* Seeing shootings much less common for white respondents


# Appendix {background-image="img/ben-sp-fD4AwrWv4t8-unsplash-crop.jpg" background-opacity=0.3}

## References

Sampson, Kirk, & Bucci. 2022. "Cohort Profile: Project on Human Development in Chicago Neighborhoods and Its Additions (PHDCN+)." *Journal of Developmental and Life-Course Criminology* 8.

Turnbull. 1976. "The empirical distribution function with arbitrarily grouped, censored and truncated data." *Journal of the Royal Statistical Society: Series B (Methodological)* 38(3):290-295.

Anderson-Bergman. 2017. "icenReg: Regression models for interval censored data in R." *Journal of Statistical Software* 81(12):1–23.